---
title: "Introduction to Computational Text Analytics"
subtitle: "Track A — R Users · ResBaz Brisbane 2026"
author: "Martin Schweinberger (LADAL, University of Queensland)"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    highlight-style: github
    code-fold: false
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# Welcome {.unnumbered}

This notebook guides you through a hands-on introduction to computational text analysis in R using the **COOEE corpus** — a collection of letters written by early immigrants to Australia (1788–1900).

You will learn to:

1. Load and clean a corpus of plain-text files
2. Perform **concordancing** (KWIC searches)
3. Compute **collocation analysis** and visualise a collocation network
4. Run a **keyword analysis** comparing COOEE to a reference corpus

**After this workshop**, you can go deeper at [ladal.edu.au/tutorials.html#text-analytics](https://ladal.edu.au/tutorials.html#text-analytics).

---

# Getting Started

## How to Get This Notebook

1. Go to **github.com/MartinSchweinberger/intro-ta-resbaz**
2. Click the green **Code** button → **Download ZIP**, unzip to your desktop  
   *— or —*  
   Fork the repo (click **Fork** top-right), then clone your fork:

```bash
git clone https://github.com/YOUR-USERNAME/intro-ta-resbaz.git
```

3. Open the file `intro-ta-resbaz.Rproj` in RStudio — this sets the working directory automatically.
4. Open `02_r_notebook.qmd`.

::: {.callout-tip}
## Why an RProject?
An `.Rproj` file anchors your working directory to the project folder, so `here::here("data/COOEE")` always finds the data regardless of where the folder lives on your machine. This is the foundation of reproducible research.
:::

## Install Required Packages

Run this once — you can skip it on future runs if packages are already installed.

```{r}
#| label: install
#| eval: false

install.packages(c(
  "tidyverse",           # dplyr, ggplot2, stringr, readr, tidyr, purrr
  "quanteda",            # core text analysis framework
  "quanteda.textstats",  # frequency and keyness statistics
  "quanteda.textplots",  # word clouds and network plots
  "readxl",              # read_xls() for .XLS metadata
  "here",                # portable file paths
  "igraph",              # network analysis
  "ggraph",              # network visualisation
  "ggrepel"              # non-overlapping text labels
))
```

## Load Packages

```{r}
#| label: setup

library(tidyverse)             # dplyr, ggplot2, stringr, etc.
library(quanteda)              # core text analysis
library(quanteda.textstats)    # frequency, keyness
library(quanteda.textplots)    # visualisations
library(readxl)                # read_xls() for .XLS metadata
library(here)                  # portable file paths
library(igraph)                # network graphs
library(ggraph)                # network visualisation
library(ggrepel)               # non-overlapping labels
```

---

# Load the COOEE Corpus

## Read the Metadata

The COOEE spreadsheet contains one row per letter with author details, year of writing, place, and more.

```{r}
#| label: load-metadata

cooee_meta <- read_xls(here("data", "COOEE", "COOEE.XLS"), skip = 1) |>
  # rename columns to easier names
  # note: Gender and Status appear twice (author + addressee), so readxl
  # disambiguates them automatically as Gender...4, Status...7, etc.
  rename(
    id               = Nr,
    author           = Name,
    birth_year       = Birth,
    author_gender    = `Gender...4`,
    origin           = Origin,
    age              = Age,
    author_status    = `Status...7`,
    year             = `Year Writing`,
    place            = `Place Writing`,
    register         = Register,
    text_type        = TextT,
    n_words          = `# of words`,
    addressee_gender = `Gender...15`,
    addressee_status = `Status...16`
  ) |>
  # ensure year is numeric
  mutate(year = as.integer(year)) |>
  # create a broad time period variable for comparisons
  mutate(
    period = case_when(
      year < 1830 ~ "Early (1788-1829)",
      year < 1870 ~ "Mid (1830-1869)",
      TRUE        ~ "Late (1870-1900)"
    )
  )

# quick look at the metadata
glimpse(cooee_meta)
```

## Read the Text Files

Each `.txt` file in `data/COOEE/` is a single letter. We read them all, extract the file ID from the filename, and join with the metadata.

```{r}
#| label: load-texts

# list all .txt files in the COOEE folder
cooee_files <- list.files(
  path       = here("data", "COOEE"),
  pattern    = "\\.txt$",
  full.names = TRUE
)

# read each file into a tibble: one row per letter
cooee_texts <- tibble(filepath = cooee_files) |>
  mutate(
    # extract the letter ID from the filename (e.g., "1-001" from "1-001.txt")
    id   = str_remove(basename(filepath), "\\.txt$"),
    # read the full text of each file as a single string
    text = map_chr(filepath, \(f) read_file(f))
  ) |>
  select(id, text)

# join text with metadata
cooee <- cooee_texts |>
  left_join(cooee_meta, by = "id")

cat("Letters loaded:", nrow(cooee), "\n")
cat("Years covered:", min(cooee$year, na.rm = TRUE), "-", max(cooee$year, na.rm = TRUE), "\n")
```

---

# Clean the Corpus

Before analysing text, we need to remove noise. In historical letters this includes: header/footer lines added during digitisation, XML or markup tags, and non-linguistic sequences.

::: {.callout-important}
## Pre-processing decisions matter
Every cleaning step is a methodological choice that affects results. Always document what you remove and why.
:::

## Basic Cleaning Pipeline

```{r}
#| label: clean-texts

cooee_clean <- cooee |>
  mutate(
    text_clean = text |>
      # remove any XML/HTML tags that may appear in digitised texts
      str_remove_all("<[^>]+>") |>
      # remove lines that begin with common header markers
      str_remove_all("(?m)^(Source|Pages|Transcribed|\\[)[^\n]*\n?") |>
      # collapse multiple whitespace characters to a single space
      str_replace_all("\\s+", " ") |>
      # remove leading and trailing whitespace
      str_trim()
  )

# compare a sample before and after cleaning
cat("--- BEFORE CLEANING ---\n")
cat(str_sub(cooee_clean$text[1], 1, 300), "\n\n")

cat("--- AFTER CLEANING ---\n")
cat(str_sub(cooee_clean$text_clean[1], 1, 300), "\n")
```

::: {.callout-note}
## What `str_remove_all` and `str_replace_all` do
- `str_remove_all(pattern)` — finds every occurrence of `pattern` and deletes it  
- `str_replace_all(pattern, replacement)` — finds every occurrence and swaps it for `replacement`  
- `\\s+` is a regular expression meaning "one or more whitespace characters"  
:::

---

## 🔧 Task 1: Extend the Cleaning Pipeline {.unnumbered}

Look at the raw text of a few letters. Find one type of noise or unwanted pattern not handled above and add a cleaning step.

```{r}
#| label: task1
#| eval: false

# Inspect the raw text of letter number 5 to spot patterns
cat(str_sub(cooee_clean$text[5], 1, 500))

# YOUR TURN: add a cleaning step below and explain what it removes
cooee_clean <- cooee_clean |>
  mutate(
    text_clean = text_clean |>
      str_remove_all("YOUR PATTERN HERE")  # describe what this removes
  )
```

---

# Build a quanteda Corpus

`quanteda` works with three objects: a **corpus** (texts + metadata), **tokens** (tokenised text), and a **document-feature matrix** (counts).

```{r}
#| label: build-corpus

# build the quanteda corpus object
cooee_corpus <- corpus(
  cooee_clean,
  docid_field = "id",
  text_field  = "text_clean"
)

# attach metadata as document variables
docvars(cooee_corpus, "year")   <- cooee_clean$year
docvars(cooee_corpus, "period") <- cooee_clean$period
docvars(cooee_corpus, "gender") <- cooee_clean$author_gender
docvars(cooee_corpus, "origin") <- cooee_clean$origin

# summary of the corpus
summary(cooee_corpus, n = 5)
```

---

# Concordancing

Concordancing (KWIC = Keyword In Context) shows every occurrence of a search term with its surrounding context.

## Basic KWIC Search

```{r}
#| label: concordance-basic

# tokenise the corpus
cooee_tokens <- cooee_corpus |>
  tokens(
    remove_punct   = TRUE,
    remove_numbers = TRUE,
    remove_symbols = TRUE
  ) |>
  tokens_tolower()

# search for "land" with 6 words of context on each side
land_kwic <- kwic(
  cooee_tokens,
  pattern = "land",
  window  = 6
)

# display the first 20 hits
head(land_kwic, 20)
```

## Search With a Phrase

```{r}
#| label: concordance-phrase

# phrase() searches for multi-word sequences
free_land_kwic <- kwic(
  cooee_tokens,
  pattern = phrase("free land"),
  window  = 5
)

head(free_land_kwic, 15)
```

## Search With a Wildcard

```{r}
#| label: concordance-pattern

# "*" is a glob wildcard: "settler*" matches settler, settlers, settlement, etc.
settler_kwic <- kwic(
  cooee_tokens,
  pattern = "settler*",
  window  = 5
)

cat("Hits for 'settler*':", nrow(settler_kwic), "\n")
head(settler_kwic, 15)
```

---

## 🔧 Task 2: Concordance Your Own Term {.unnumbered}

Choose a word you would expect to be interesting in letters from early immigrants to Australia — something related to landscape, hardship, family, or colonial life.

```{r}
#| label: task2
#| eval: false

# YOUR TURN: replace "YOUR_WORD" with your chosen search term
my_kwic <- kwic(
  cooee_tokens,
  pattern = "YOUR_WORD",
  window  = 6
)

head(my_kwic, 20)

# REFLECTION QUESTIONS:
# 1. What contexts does your word appear in?
# 2. Does the meaning change depending on the surrounding words?
# 3. Are there patterns in who uses this word (check the docnames)?
```

---

# Collocation Analysis

Collocations are words that co-occur together **more often than chance**. The key idea is not just that two words appear near each other, but that they appear together *significantly more* than you would expect given how common each word is on its own.

We compute this using a **co-occurrence matrix** and then calculate association measures (MI, t-score, log-likelihood). This is the approach used in the LADAL tutorials.

## Step 1: Build the Co-occurrence Matrix

```{r}
#| label: cooccurrence-matrix

# we work with tokens with stopwords removed
cooee_tokens_clean <- cooee_tokens |>
  tokens_remove(stopwords("en"))

# build a feature co-occurrence matrix (FCM)
# this counts how often each pair of words appears within a window of 5 tokens
cooee_fcm <- cooee_tokens_clean |>
  fcm(
    context = "window",
    window  = 5,           # look 5 tokens either side
    tri     = FALSE        # count both directions (symmetric)
  )

# how large is the matrix?
cat("FCM dimensions:", nrow(cooee_fcm), "x", ncol(cooee_fcm), "\n")
```

## Step 2: Extract Collocates of a Target Word

We focus on a single target word and find its strongest collocates using Mutual Information (MI).

```{r}
#| label: collocates-target

# choose a target word
target_word <- "land"

# extract the row for the target word from the FCM
# this gives us how often every other word co-occurs with "land"
target_cooccur <- cooee_fcm[target_word, ] |>
  # convert to a tidy data frame
  as.matrix() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("word") |>
  rename(cooccur_count = 2) |>
  # remove the target word itself
  filter(word != target_word, cooccur_count > 0) |>
  as_tibble()

# get total frequency of each word in the corpus (needed for MI)
word_freq <- cooee_tokens_clean |>
  dfm() |>
  featfreq() |>
  as_tibble(rownames = "word") |>
  rename(freq = value)

# get total token count (needed for MI calculation)
total_tokens <- sum(word_freq$freq)

# get frequency of target word
target_freq <- word_freq |>
  filter(word == target_word) |>
  pull(freq)

# compute association measures
collocates <- target_cooccur |>
  left_join(word_freq, by = "word") |>
  filter(!is.na(freq)) |>
  mutate(
    # Mutual Information: measures how much more often the pair co-occurs
    # than expected by chance (high MI = strong but possibly rare association)
    MI = log2(
      (cooccur_count * total_tokens) /
      (target_freq * freq)
    ),
    # t-score: favours frequent and reliable collocations
    t_score = (cooccur_count - (target_freq * freq / total_tokens)) /
              sqrt(cooccur_count)
  ) |>
  # keep only words that co-occur at least 3 times
  filter(cooccur_count >= 3) |>
  arrange(desc(MI))

# view the top collocates
cat("Top collocates of '", target_word, "' by Mutual Information:\n", sep = "")
head(collocates, 20)
```

::: {.callout-note}
## Why Mutual Information, not just frequency?
A word like *"the"* might co-occur with *"land"* many times, but only because *"the"* is extremely common everywhere — it tells us nothing specific about *"land"*. MI corrects for this by comparing observed co-occurrence to what we would *expect* given each word's overall frequency. High MI means the two words attract each other specifically.
:::

## Step 3: Visualise a Collocation Network

A collocation network shows which words attract each other. Nodes = words, edges = strong co-occurrence.

```{r}
#| label: collocation-network
#| fig-width: 9
#| fig-height: 7

# select the top collocates of our target word for the network
# we use the top 15 by MI with at least 3 co-occurrences
top_collocates <- collocates |>
  filter(MI > 2) |>
  slice_head(n = 15)

# build an edge list: target word connects to each collocate
edges <- top_collocates |>
  mutate(from = target_word) |>
  select(from, to = word, weight = MI)

# build the igraph object
collocation_graph <- graph_from_data_frame(edges, directed = FALSE)

# plot with ggraph
ggraph(collocation_graph, layout = "fr") +
  # edges: width and transparency proportional to MI strength
  geom_edge_link(
    aes(width = weight, alpha = weight),
    colour      = "#51247A",
    show.legend = FALSE
  ) +
  # nodes
  geom_node_point(
    aes(size = ifelse(name == target_word, 8, 4)),
    colour      = "#00A2C7",
    show.legend = FALSE
  ) +
  # word labels
  geom_node_text(
    aes(label = name),
    repel    = TRUE,
    size     = 3.5,
    fontface = "bold"
  ) +
  scale_edge_width(range = c(0.4, 2.5)) +
  scale_edge_alpha(range = c(0.3, 0.9)) +
  labs(
    title    = paste0("Collocates of '", target_word, "' — COOEE Corpus"),
    subtitle = "Edges connect words with high Mutual Information (MI > 2, min. 3 co-occurrences)",
    caption  = "Source: COOEE corpus · Analysis: quanteda + ggraph"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold", margin = margin(b = 4)),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    plot.caption  = element_text(size = 8,  color = "grey60")
  )
```

---

## 🔧 Task 3: Explore Collocations {.unnumbered}

```{r}
#| label: task3
#| eval: false

# QUESTION 1: Look at the top 10 collocates of "land" by MI.
# What do they tell you about how "land" was discussed in these letters?

# QUESTION 2: Change the target word and re-run the analysis.
# Try: "home", "ship", "work", "wife", or "family"
target_word <- "home"   # change this

# Re-run the co-occurrence extraction and network plot above.
# How does the collocation profile of your new word differ from "land"?

# QUESTION 3: Compare MI vs t-score rankings.
# Run: head(arrange(collocates, desc(t_score)), 10)
# Which words appear in the t-score top 10 but not the MI top 10?
# Why might that be?
head(arrange(collocates, desc(t_score)), 10)
```

---

# Keyword Analysis

Keyword analysis identifies words that are statistically *distinctive* in the COOEE corpus compared to a reference corpus (BROWN). A keyword is not simply a frequent word — it is a word that appears significantly more often than you would expect based on general English usage.

## Load the BROWN Reference Corpus

```{r}
#| label: load-brown

# note: BROWN files use upper-case extension .TXT
brown_files <- list.files(
  path       = here("data", "BROWN"),
  pattern    = "\\.TXT$",
  full.names = TRUE
)

brown_texts <- tibble(filepath = brown_files) |>
  mutate(
    id   = str_remove(basename(filepath), "\\.TXT$"),
    text = map_chr(filepath, \(f) read_file(f)) |>
      # extract only words (letter sequences), discard all numbers/codes/punctuation
      str_extract_all("[a-zA-Z]+") |>
      map_chr(\(words) paste(words, collapse = " "))
  ) |>
  select(id, text)  |>
  # remove sngle letters
  dplyr::mutate(text = stringr::str_remove_all(text, "\\b\\w\\b"))

cat("BROWN files loaded:", nrow(brown_texts), "\n")

# inspect
head(brown_texts)

```

## Build a Combined DFM

For keyness analysis we need a **Document-Feature Matrix** with both corpora labelled. The key is to embed the `corpus_label` variable in the data *before* building the corpus, so it is available as a docvar for grouping.

```{r}
#| label: keyness-dfm

# combine COOEE and BROWN into one labelled tibble
combined_texts <- bind_rows(
  cooee_clean |>
    select(id, text = text_clean) |>
    mutate(corpus_label = "COOEE"),
  brown_texts |>
    mutate(corpus_label = "BROWN")
)

# build corpus from the combined tibble
combined_corpus <- corpus(
  combined_texts,
  docid_field = "id",
  text_field  = "text"
)

# assign corpus_label explicitly as a docvar
# (safest approach: pull directly from the source tibble)
docvars(combined_corpus, "corpus_label") <- combined_texts$corpus_label

# build the DFM, then group by corpus_label
# grouping collapses all COOEE documents into one row and all BROWN into one row
combined_dfm <- combined_corpus |>
  tokens(
    remove_punct   = TRUE,
    remove_numbers = TRUE,
    remove_symbols = TRUE
  ) |>
  tokens_tolower() |>
  tokens_remove(stopwords("en")) |>
  dfm() |>
  dfm_group(groups = docvars(combined_corpus, "corpus_label"))

# confirm we have exactly 2 documents (COOEE and BROWN)
combined_dfm
```

## Compute Keyness

```{r}
#| label: keyness-compute

# compute keyness: COOEE is the target, BROWN is the reference
# lr = log-likelihood ratio (G2) — the most robust measure
cooee_keyness <- textstat_keyness(
  combined_dfm,
  target  = "COOEE",
  measure = "lr"
)

# top 20 keywords (most distinctive in COOEE vs BROWN)
head(cooee_keyness, 20)
```

## Visualise Keywords

```{r}
#| label: keyness-plot
#| fig-width: 9
#| fig-height: 6

# plot top 15 positive keywords (more in COOEE) and top 15 negative (more in BROWN)
textplot_keyness(
  cooee_keyness,
  n         = 15,
  color     = c("#51247A", "#00A2C7"),
  labelsize = 3
) +
  labs(
    title    = "Keyword Analysis: COOEE vs. BROWN Corpus",
    subtitle = "Purple = distinctive in COOEE letters · Teal = distinctive in BROWN",
    caption  = "Measure: log-likelihood ratio (G2)"
  ) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40", size = 10)
  )
```

---

## 🔧 Task 4: Interpret the Keywords {.unnumbered}

```{r}
#| label: task4
#| eval: false

# QUESTION 1: Look at the top 10 keywords for COOEE.
# What topics or themes do these words suggest?
# What does this tell us about what was important to early immigrants?

# QUESTION 2: Look at the top words that are MORE common in BROWN.
# Why might these be more common in standard American English
# than in 19th-century Australian immigrant letters?

# QUESTION 3: Try a different keyness measure.
# Change measure = "lr" to measure = "chi2"
cooee_keyness_chi2 <- textstat_keyness(
  combined_dfm,
  target  = "COOEE",
  measure = "chi2"
)
head(cooee_keyness_chi2, 15)

# Compare to the log-likelihood results.
# Which words appear in both lists? Which differ?
```

---

## 🔧 Task 5: Diachronic Keyword Analysis {.unnumbered}

Can we detect how language changed across the three time periods in COOEE?

```{r}
#| label: task5
#| eval: false

# Build a DFM grouped by time period
cooee_period_dfm <- cooee_corpus |>
  tokens(remove_punct = TRUE, remove_numbers = TRUE) |>
  tokens_tolower() |>
  tokens_remove(stopwords("en")) |>
  dfm() |>
  dfm_group(groups = "period")

cooee_period_dfm

# YOUR TURN: compute keyness comparing Late to Early letters
late_vs_early <- textstat_keyness(
  cooee_period_dfm,
  target = "Late (1870-1900)"
)

head(late_vs_early, 15)

# Plot the results
textplot_keyness(late_vs_early, n = 20)

# REFLECTION: What words become more distinctive in later letters?
# What words characterise earlier letters?
# What historical changes might explain these differences?
```

---

# Where to Go Next

| Method | LADAL Tutorial |
|---|---|
| Concordancing | [ladal.edu.au/tutorials/concordancing](https://ladal.edu.au/tutorials/concordancing/concordancing.html) |
| Collocation Analysis | [ladal.edu.au/tutorials/collocations](https://ladal.edu.au/tutorials/collocations/collocations.html) |
| Keyword Analysis | [ladal.edu.au/tutorials/keywords](https://ladal.edu.au/tutorials/keywords/keywords.html) |
| Sentiment Analysis | [ladal.edu.au/tutorials/sentiment](https://ladal.edu.au/tutorials/sentiment/sentiment.html) |
| Topic Modelling | [ladal.edu.au/tutorials/topic](https://ladal.edu.au/tutorials/topic/topic.html) |
| POS Tagging | [ladal.edu.au/tutorials/pos_tagging](https://ladal.edu.au/tutorials/pos_tagging/pos_tagging.html) |
| Network Analysis | [ladal.edu.au/tutorials/network_analysis](https://ladal.edu.au/tutorials/network_analysis/network_analysis.html) |

::: {.callout-tip}
## Corpus Linguistics Showcase
For a complete worked example using corpus methods on historical text (including COOEE!), see:  
[Showcase: Comparing Methods for Conceptual Maps](https://ladal.edu.au/tutorials/conceptual_maps_comparison/conceptual_maps_comparison.html)
:::

---

# Session Info {.unnumbered}

```{r}
#| label: session-info
sessionInfo()
```
